---
title: "FY24 and FY25 CAPS"
author: "IFPTE 8a"
format: html
editor: visual
execute:
  dir: project
  cache: true
  echo: false
---

## Background

IFPTE 8a requested data on FY2024 and FY2025 CAPS scores, bonuses, RIF credits and performance pay increases from management to inform how recent changes have impacted scoring and compensation for bargaining unit members. This document summarizes the data we received.

```{r}
#| message: false
#| warning: false
library(readxl)
library(tidyverse)
library(janitor)
library(ggridges)
library(ggrepel)
library(here)
library(knitr)
```

## Data import and cleaning

We have anonymized data for all bargaining unit members from the AFSC and NWFSC from FY2024 and FY2025. The AFSC data includes path, band, score, performance pay increase (percent), bonus, and RIF credits (FY25 only). The NWFSC data include path, band score, performance pay increase (absolute numbers), bonus, RIF credits, and pool. For the NWFSC, the pool for ZA and ZP is ZA/ZP (still unclear what this means).

```{r}
#| message: false
#| warning: false
#Read in AK data
#note: no info on RIF credits for AK 2024
#note: no "pool" column
ak24 <- read_xlsx(here('data', 'FY24 and FY25 AKC CAPS Information for IFPTE Request.xlsm'), sheet = "AKC FY24") |>
  clean_names() |>
  mutate(year = "FY24") |>
  mutate(center = "AFSC") |> 
  mutate(percent_increase = percent_increase*100) #this is getting read in as a proportion, quick fix

ak25 <- read_xlsx(here('data','FY24 and FY25 AKC CAPS Information for IFPTE Request.xlsm'), sheet = "AKC FY25") |>
  clean_names() |>
  mutate(year = "FY25") |>
  rename(rif = rif_credit) |>
  mutate(center = "AFSC")

#read in NW data
nw24 <- read_xlsx(here("data", "NWFSC FY24 FY25 CAPS data Union request April 2026.xlsx"), sheet = "FY24")|>
  clean_names() |>
  mutate(year = "FY24") |>
  rename(score = fy24_score,
         rif = rif_credit) |>
  mutate(center = "NWFSC")

nw25 <- read_xlsx(here("data", "NWFSC FY24 FY25 CAPS data Union request April 2026.xlsx"), sheet = "FY25")|>
  clean_names() |>
  mutate(year = "FY25") |>
  rename(score = fy25_score) |>
  mutate(center = "NWFSC")

#Combine all together 
all <- bind_rows(nw24, nw25, ak24, ak25) |>
  select(-x1) |>
  mutate(percent_increase = ifelse(is.na(percent_increase) & center == "AFSC",
                                   0,
                                   percent_increase),
         performance_pay_increase = ifelse(is.na(performance_pay_increase) & center == "NWFSC",
                                           0,
                                           performance_pay_increase)) |> #assuming blank means no increase at AFSC
  mutate(pool = ifelse(is.na(pool), 
                       path,  
                       pool)) |> #assuming pool is path at AFSC, but we should check
  mutate(in_top_30 = ifelse(rif == 10, 1, 0)) |> #categorical variable for whether score was in the top 30th based on RIF credits
  #Also keep separate AK/NW data
  mutate(
    year = factor(year, levels = c("FY24", "FY25")),
    path = factor(path),
    year_top30 = interaction(year, in_top_30, drop = TRUE)
  ) |> 
  #how many people do we have in these categories?
  group_by(center, year, path) |> 
  mutate(n_in_path = n()) |> 
  group_by(center, year, pool) |> 
  mutate(n_in_pool = n()) |> 
  ungroup()

nw <- filter(all, center == "NWFSC") |>
  mutate(got_raise = ifelse(performance_pay_increase > 0, 1, 0))

ak <- filter(all, center == "AFSC") |>
  mutate(got_raise = ifelse(percent_increase > 0, 1, 0))
```

## Analysis

We did a brief comparison of the distribution of scores, the scores associated with 5 or 10 RIF credits (which should correspond to being in the lower 70th or upper 30th percentile of the pay pool), and the relationship between score and performance increase. For all analyses, any year/center/path combination with fewer than three employees are not shown to avoid inadvertently revealing any identifying information. Table 1 shows a summary of the information provided by year, center, and path.

```{r}
#summary stats by year/center/path
summary_by_center_year_path <- all |> 
  filter(n_in_path >= 3) |> 
  group_by(year, center, path) |>
  summarise(#quant70 = quantile(score, 0.7),
    mean_score = round(mean(score),1),
    med_score = median(score),
    mean_bonus = round(mean(bonus),1),
    med_bonus = median(bonus),
    mean_absolute_inc = mean(performance_pay_increase),
    med_absolute_inc = median(performance_pay_increase),
    mean_pct_inc = round(mean(percent_increase),1),
    med_pct_inc = median(percent_increase),
    total_bonus = sum(bonus),
    total_inc = sum(performance_pay_increase),
    pct_5rif = round(sum(rif == 5)/n(),2)*100,
    pct_10rif = round(sum(rif == 10)/n(),2)*100, 
    .groups = "drop_last") |>
  arrange(center, path, year) |> 
  as.data.frame()

summary_by_year_center <- all |> 
  group_by(year, center) |> 
  summarise(total_bonus = sum(bonus),
            total_increase = sum(performance_pay_increase),
            pct_10_rif = round(sum(rif == 10)/n(),2)*100,
            mean_pct_increase = round(mean(percent_increase),1),
            med_pct_increase = round(median(percent_increase),1),
            .groups = "drop_last")

```

```{r}
#| label: tbl-summary
#| tbl-cap: "Summary of bonuses, performance pay increases, and RIF credits by path, year, and center"
kable(summary_by_center_year_path)
```

At the NWFSC, bonuses totaled \$`r format(subset(summary_by_year_center, center == "NWFSC" & year == "FY24")$total_bonus, big.mark=",")` in FY24 and \$`r format(subset(summary_by_year_center, center == "NWFSC" & year == "FY25")$total_bonus, big.mark=",")` in FY25. At the AFSC, bonuses totaled \$`r format(subset(summary_by_year_center, center == "AFSC" & year == "FY24")$total_bonus, big.mark=",")` in FY24 and \$`r format(subset(summary_by_year_center, center == "AFSC" & year == "FY25")$total_bonus, big.mark=",")` in FY25. At the NWFSC, performance pay increases totaled \$`r format(subset(summary_by_year_center, center == "NWFSC" & year == "FY24")$total_increase, big.mark=",")` in FY24 and \$`r format(subset(summary_by_year_center, center == "NWFSC" & year == "FY25")$total_increase, big.mark=",")` in FY25. The AFSC provided performance pay data in percents, and the median performance pay increase was `r format(subset(summary_by_year_center, center == "AFSC" & year == "FY24")$med_pct_increase, big.mark=",")`% in FY2024 and `r format(subset(summary_by_year_center, center == "AFSC" & year == "FY25")$med_pct_increase, big.mark=",")`% in FY2025. At the NWFSC, `r format(subset(summary_by_year_center, center == "NWFSC" & year == "FY24")$pct_10_rif, big.mark=",")`% of bargaining unit employees were given 10 RIF credits in both FY24 and FY25, corresponding to the percent of employees who were in the top 30th percentile of their respective pay pools. At the AFSC, this number was `r format(subset(summary_by_year_center, center == "AFSC" & year == "FY25")$pct_10_rif, big.mark=",")`% in FY25 (data for FY24 not provided).

We used the maximum scores with 5 RIF credits and the minimum scores with 10 RIF credits to estimate the scores associated with the 70th/30th percentile of scores in a pay pool. Figure 1 shows these scores for each path, center, and year.

```{r}
#| label: fig-rif_cutoff
#| fig-cap: "Maximum score with 5 RIF credits (assumed lower 70th percentile) and minimum score with 10 RIF credits (assumed upper 30th percentile)"
rif_cutoff <- all |> 
  filter(n_in_path >= 3) |> 
  group_by(year, center, path) |> 
  summarise(
    highest_5 = max(score[rif == 5]),
    lowest_10 = min(score[rif == 10]),
    .groups = "drop"
  )
#Potentially something a bit funky with NWFSC ZA/ZP rif credits in 2024 IF they should be considered the same pool

ggplot(rif_cutoff |> filter(!is.na(highest_5)), aes(y = factor(year))) +
  # Draw the connecting gap segment
  geom_segment(aes(x = highest_5, xend = lowest_10, yend = factor(year)), 
               color = "grey60", linewidth = 1) +
  # Point for highest 5 RIF
  geom_point(aes(x = highest_5, color = "Highest RIF 5 score"), size = 3) +
  # Point for lowest 10 RIF
  geom_point(aes(x = lowest_10, color = "Lowest RIF 10 score"), size = 3) +
  facet_grid(center ~ path) +
  scale_color_manual(values = c("Highest RIF 5 score" = "blue", "Lowest RIF 10 score" = "red")) +
  labs(
    title = "Estimated 70-30 Cutoff",
    x = "Score",
    y = "Year",
    color = ""
  ) +
  geom_text_repel(
    aes(x = highest_5, label = round(highest_5, 1)),
    nudge_x = -0.5,
    direction = "x",
    size = 3.2,
    segment.color = NA 
  ) +
  
  # Repelled Labels for RIF 10 (pushed right)
  geom_text_repel(
    aes(x = lowest_10, label = round(lowest_10, 1)),
    nudge_x = 0.5,
    direction = "x",
    size = 3.2,
    segment.color = NA
  )+
  theme_bw()
```

The overall distribution of scores is shown in Figure 2.

```{r}
#| message: false
#| warning: false
#| label: fig-score-ridgeplot
#| fig-cap: "Density plot of scores by year, center, and path. Vertical lines represent quantiles."

ggplot(all |>  filter(n_in_path >= 3), aes(x = score, y = year))+
  geom_density_ridges2(quantile_lines =TRUE)+
  facet_grid(path~center)+
  theme_bw()+
  ggtitle("Distribution of scores")

```

The smoothed relationship between score and performance pay increase is shown in Figures 3 and 4. Individual scores are not shown in order to avoid inadvertently revealing personal information. Note that we do not have information on which employees are at the top of their pay band ("capped out").

```{r}
#| message: false
#| warning: false
#| label: fig-nw_score_perf_pay
#| fig-cap: "Smoothed plot of score versus performance pay increase (NWFSC)."
ggplot(nw, aes(x = score, y = performance_pay_increase, color = year, group = year))+
  geom_smooth()+
  ylab("Performance pay increase (dollars)")+
  #facet_wrap(~path)+
  theme_bw()
```

```{r}
#| message: false
#| warning: false
#| label: fig-ak_score_perf_pay
#| fig-cap: "Smoothed plot of score versus performance pay increase (AFSC)."
ggplot(ak, aes(x = score, y = percent_increase, color = year, group = year))+
  geom_smooth()+
  #facet_wrap(~path)+
  ylab("Performance pay increase (%)")+
  theme_bw()
```

## Additional context

Variation in how bonuses are doled out, leveling, etc.
